A Privacy Aware Mobile Sensor Application

### 2015-02-01 paper.tex

\title{Privacy Aware Mobile Sensor Applications}
\author{Heinrich Hartmann \and Tim Wambach \and Maximilian Meffert \and R\"udiger Grimm}
\institute{University of Koblenz-Landau}
\maketitle

@@

In this article we analyze the privacy aspects of a mobile sensor
application used for recording urban travel patterns as part of a
travel-survey service. This service has been developed and
field-tested within the Live+Gov EU Project. The privacy analysis
follows a structured approach established in
\cite{Grimm:ItSecRefModel}. Eight privacy recommendations are derived,
and have already led to corresponding enhancements of the
travel-survey service.

@@

The rise of mobile smart-phones equipped with a wide range of sensors
and the abundance mobile internet connectivity has paved the way to a
whole new generation of mobile services which ease the daily life of
citizens of the western developed world. Apps like Google
Maps\footnote{\url{http://google.com/maps}} allows the user to
navigate effectively in unknown cities. Others
(e.g. Strava\footnote{Citation Needed}) track sports activities in
order to optimize training plans or engage in virtual competitions .

Besides the immediate benefits citizens using these services reveal
a wealth of which can be used to draw a very detailed picture of the
private life of the citizen. For instance, GPS location tracking can
be used to infer shopping habits or associations with political groups
(when meetings are attended). Accelerometer data can be used to detect
medical conditions like walking disabilities. All this data is highly
sensitive to the citizens privacy and can be used against the citizen
if it falls into the wrong hands.

Providers of these new services are faced with a fundamental trade-off
between features and convenience of the service and the protection of
the citizen's privacy. The service has to collect enough data to
support the basic service-promises. If too much data is collected the
privacy of the citizen is put at risk, with immediate implications for
the customer trust relation and the acceptance of the service. Also
leaks of private data can jeopardize the whole business.

@@

%\TODO{MM:\\
%- Shorten\\
%- DONE - Remove Implicit Type Violations\\
%- Update figure\\
%- RG: Discuss alternative definitions of privacy\\
%}

\noindent
Defining privacy is a challenge which seems impossible. This is well put to words by Serge Gutwirth, who notes:
\begin{quote}
The notion of privacy remains out of the grasp of every academic chasing it. Even when it is cornered by such additional modifiers as `our' privacy, it still finds a way to remain elusive. \cite{Gutwirth}
\end{quote}
In this section we introduce the concept of privacy as control over information.
We specify types of information we have to deal with in the context of privacy.
And finally we outline the impact mobile sensors can have on one's privacy.

\subsection{Privacy as Control over Information}
%We may need to remove the following paragraph depinding on RG's alternative definition of privacy
%Many definitions seem to focus on the vulnerabilities of privacy.
%Thus a great amount of work is invested in finding threats to prohibit instead of describing why privacy is so valuable to us.
%This way around one could derive measures to ensure and secure its value. \cite{7ToP}

%One particular scholar, who did not follow this way of thinking is Charles Fried.
One possible definition of privacy is given by Charles Fried in his work \textit{An Anatomy of Values}\cite{CFried:Privacy}.
He questioned, why we are intuitively sensitive to privacy violations.
But he did not assert privacy as an intrinsic value by itself, he rather stated:
\begin{quote}
\textbf{Privacy is} not simply an absence of information about us in the minds of others;
rather it is \textbf{the control we have over information about ourselves}. \cite{CFried:Privacy}
\end{quote}
According to Fried privacy is our ability to create and modulate the relationships to other humans.
Which makes us intuitively sensitive for its violation, because this affects what defines our identity.
Also, the ability to build and maintain relationships is essential to society.
This is what makes privacy valuable and deserving of protection. 
\cite{sep-privacy}\cite{CFried:Privacy}

\subsubsection{Relationship Creation.}

% Privacy + Trust ~> Relationship
% ===============================
% - A shares intimate information with B
% - A trusts B not to reveal that information by not constantly monitoring B, i.e. respecting B's privacy
% - B has now the possibility to share intimate information with A

We share information of great intimacy only with persons we deem trustworthy. 
Moreover, we trust those persons to not reveal information about us to others by not constantly monitoring them, i.e. respecting their privacy.
Trust needs the possibility of unknown failure.
If we would constantly monitor our partners, they cannot fail unnoticed nor can they willingly share intimate information with us.
By trusting them, we create the possibility for them to share intimate information with us.
\cite{CFried:Privacy}

\subsubsection{Relationship Modulation.}
Fried argues, depending on the person we are interacting with, we are changing the degree of intimacy we share.
If we are to share private information with others, we are most likely aware that this action is privacy related.
In that case we are able to selectively disclose information along the two dimensions of quantity and quality.
\cite{CFried:Privacy}

\subsubsection{Alternative Definitions.}
Using Fried's anatomy of privacy for this analysis is suitable because of two points.
Despite the fact his text was published in 1970, he already included technologies to its viewpoint that did not only monitor location, but also record biometric data \cite{CFried:Privacy}. 
Secondly, the concept of control over information does not enforce data minimization.
Fried's definition complies with what we call \textit{self data protection}, where persons communicate their data self-determined.
This is the opposite of \textit{system data protection}, where the collection of certain data is prohibited.

\TODO{RG: Discuss alternative definitions of privacy}

%=======================================================================================

\begin{LGContent-Hidden}
Defining privacy is a challenge which seems impossible. This is well put to words by Serge Gutwirth, who notes:

\begin{quote}
The notion of privacy remains out of the grasp of every academic chasing it. Even when it is cornered by such additional modifiers as `our' privacy, it still finds a way to remain elusive. \cite{Gutwirth}
\end{quote}

%Many researchers seem to only ``focus on the ways in which privacy can be infringed'' \cite{7ToP}.
%Thus they try to create taxonomies of \emph{privacy harms} instead of taxonomies of \emph{privacy types}.
%Those two differ in the respect that the former focuses on threats to prohibit whereas the latter focuses on values to protect.
%So one should rather evaluate what aspects are precious about privacy and develop measures to ensure their security than only forbid single actions against it \cite{7ToP}.

Many researchers seem to only ``focus on the ways in which privacy can be infringed'' \cite{7ToP}.
Thus they invest a great amount of work in defining threats to prohibit instead of describing why privacy is so valuable to us.
This way around one could derive measures to ensure and secure its value \cite{7ToP}.
One particular scholar who does this in the context of digital monitoring is Charles Fried.
He questioned, why we are intuitively sensitive to violations of privacy.
But he did not assert privacy as an intrinsic value by itself, he rather sated:

\begin{quote}
Privacy is not simply an absence of information about us in the minds of others;
rather it is \textbf{the control we have over information about ourselves}. \cite{CFried:Privacy}
\end{quote}

According to Fried privacy is a \textit{rational context}.
Which means that one is aware of such context's existence during a rational action.
If we are to share private information with others, we are most likely aware that this action is privacy related.
In that case we are able to selectively disclose information along the two dimensions of quantity and quality.
This leads Fried to his thesis, that privacy is one's ability to create and modulate his or hers social relationships, namely: friendship, love and trust. \cite{CFried:Privacy}

Depending on the conversation partner, we change the degree of intimate information we share if it is a total stranger, colleague, close friend or lover.
With close friends or lovers we share information of great intimacy we do not share with anyone else.
Moreover, we trust those persons to not reveal information about us to others by respecting their privacy.
Trust needs the possibility of unknown failure.
If we would constantly monitor our partners, they cannot fail unnoticed nor can they willingly share that information with us.
Thus they could not trust us anymore. \cite{CFried:Privacy}

So privacy or the its possibility, according to Fried, is the foundation of our core relations: friendship, love and trust.
And thus it is valuable, because those relations are essential to human society \cite{CFried:Privacy}.
Using his anatomy of privacy as foundation of our analysis is suitable because of two points.
At first, Fried's study on the understanding of privacy provided a great contribution to the research on the same term in philosophy \cite{sep-privacy} and computer science \cite{SociotechnicalArchitectureForOnlinePrivacy}.
Secondly, despite the fact his text was published in 1970, he already included technologies to its viewpoint that did not only monitor location, but also record biometric data.
\end{LGContent-Hidden}

@@

The first step is the \emph{world view} where all components are described in their current state. It consists of the following components:
\begin{itemize}

	\item \textbf{Assets.} 
	Things of value to one or more stakeholders.

	\item \textbf{IT-Systems.}
	The relevant IT-Systems under study. Already described in Section \ref{subsubsection:it-systems}.

	\item \textbf{Actors.}
	All identifiable stakeholders of the system under study.

	\item \textbf{Conflicts of Interests.}
	Different actors have different interests which can be in conflict which each other.

	\item \textbf{Vulnerabilities.}
	All identifiable weaknesses in the IT-System e.g. a weak passwords.

	\item \textbf{Interactions.}
	This point captures all possible interactions between assets, IT-Systems, humans, and vulnerabilites as partly described in Section \ref{subsubsection:it-systems} and will be further analyzed in \ref{subsec:pot-analysis}.

\end{itemize}

@@

In this scenario we focus our attention to only one asset: The privacy of the citizen.
Our definition of privacy is described in detail in Section \ref{sec:privacydef} and introduced in Section \ref{sec:privacytypes} the seven different types of privacy.

@@

\subsection{Step 2. Potential Analysis}
\label{subsec:pot-analysis}

The \emph{Potential Analysis} (or \emph{Potential View}) displays the intended and unintended interactions of the components in the world view.
The intended interactions support the underlying business objectives.
Unintended interactions lead to threats. The potential view consists of the following components:

\begin{itemize}

	\item \textbf{Business Objectives.}
	Interaction of IT-system and actors that realize a business goals of the system owner were already described in Section \ref{subsubsection:it-systems} and \ref{subsec:world-analysis}.

	\item \textbf{Threats.}
	A threat is a potential interaction that destroys or harms assets of the system.
	Concrete realizations of threats can be \emph{attacks} or \emph{accidents}.
	Attacks are executed by an actor in response to a conflict of interest.
	Accidents are harmful interactions that are not willfully caused by an actor.

	\item \textbf{Chances/Risk.}
	Evaluation of chances and risks associated to the business objectives and threats. The risk associated to a threat is its expected loss.

	\item \textbf{Security Requirements.}
	Interactions (e.g. threats) that shall not occur within the system in order to achieve its business objectives.  We included these security requirements into our analysis in Section \ref{subsec:priv-rec}.

\end{itemize}

\subsubsection{Threat Specification}

A threat is a potential interaction of the components that targets an asset.
We restrict ourselves to the case of attacks, and the asset of privacy.
An attack is an interaction that is executed by an actor in response to a conflict of interest by exploiting a vulnerability of the system.
The alternative interaction of accidents are less relevant for us, since the violation of privacy always requires an actor that takes advantage of personal data.

We describe threats for the citizen privacy in the Live+Gov system.
This list can necessarily not be complete, but we make a best effort to cover the most relevant cases.

\textbf{T1. Insufficient Control Features.}
As soon as collected data of Citizens is stored on System Provider servers, all control over that data is lost. Although, control capabilities for Citizens add to the system complexity, which could motivate to omit those. Therefore interests of Citizens and System Providers are in conflict, namely it is the Citizen's \textit{Privacy vs. System Complexity} for System Providers. This is an abstract threat provoked by a general \textit{Missing Privacy Awareness} in the minds of all actors in the Live+Gov context.

\textbf{T2. Excessive Data Mining.}
The System Provider and/or the Local Authority secretly extract more private information from the collected data, than the Citizen agreed to.
But results of the mining process create no disadvantages for Citizens, because there is no disclosure to third parties.
This could be the case for a System Provider, who wants to test a new product and uses the pre-existing data collection.
Or for a Local Authority, who wants to analyze the data collection regarding fare evasion.
However, the Citizen has not agreed to such data processing nor could he, since it is conducted secretly.
This disables a Citizen to control his data adequately.
Thus there are two possible conflicts: \textit{Privacy vs. Financial Profit of Serivce Provider} and  \textit{Privacy vs. Buisness Intelligence of Local Authority}
The threat can be provoked by either lax data handling policies of both System Providers and Local Authorities, or a weak law enforcement of existing supervision.
But the main issue, which can lead to such threats, is again a general \textit{Missing Privacy Awareness}.

>>>>

\title{A Privacy Aware Mobile Sensor Application}

\author{Heinrich Hartmann \and Tim Wambach \and Maximilian Meffert \and R\"udiger Grimm}
\institute{University of Koblenz-Landau}
\maketitle

@@

In our research work we have implemented a privacy aware mobile sensor application [diesen ersten Satz ggf. wieder streichen]. 
In this paper we outline a scenario of a mobile traffic survey and analyse its privacy risks according to a reference model for IT security analysis.
This leads to concrete recommendations for privacy protecting measures that are implemented in the system, for example a Web-based dashboard which allows the users to view, modify and delete all personal data that is stored in the system.

Vorschlag HH: In this article we analyze the privacy aspects of a mobile sensor
application used for recording urban travel patterns as part of a
travel-survey service. This service has been developed and
field-tested within the Live+Gov EU Project. The privacy analysis
follows a structured approach established in
\cite{Grimm:ItSecRefModel}. Eight privacy recommendations are derived,
and have already led to corresponding enhancements of the
travel-survey service.

\keywords{
privacy protection, IT security analysis, sensor data, mobile phones, traffic survey
}

@@

The rise of mobile smart-phones equipped with a wide range of sensors
and the abundance mobile internet connectivity has paved the way to a
whole new generation of mobile services which ease the daily life of
citizens using them. Apps like Google
Maps\footnote{\url{http://www.google.com/mobile/maps/}} allows the user to
navigate effectively in unknown cities; others track sports activities
in order to optimize training plans or engage in virtual competitions.

Besides the immediate benefits these services reveal a wealth of
private information about the citizen that capture a very detailed
picture of his life. For instance, GPS location tracking can be used
to infer shopping habits or associations with political groups (when
meetings are attended) and accelerometer data can be used to detect
medical conditions like walking disabilities. All this data is highly
sensitive to the citizens privacy and can be used against the citizen
if it falls into the wrong hands.

Providers of these new services are faced with a fundamental trade-off
between features and convenience of the service and the protection of
the citizen's privacy. The service has to collect enough data to
support the basic service promises. If too much data is collected the
privacy of the citizen is put at risk, with immediate implications for
the customer trust relation and the acceptance of the service. Also
leaks of private data can jeopardize the whole business.

@@

\subsection{Privacy and Self Data Protection}
More than one hundred years ago, 1890, the American politician Louis Brandis had specified privacy as the right to be left alone \cite{WarrenBrandeis:RightToPrivacy}. From the data processing point of view, this right is best expressed by the absence of information about a person in the mind of others. Indeed, this principle of “data minimization” is still fundamental to modern data protection legislation. The modern principles of data protection are codified by many legal systems in different countries, especially in Europe \cite{EU-directive2}, and by the US save harbor principles. However, modern data protection is more than only the absence of personal data. It is based on the personal right on self-determination over information and communication \cite{BVerfG:census}. This is, for example, very well expressed by Fried’s definition of 1970:
\begin{quote}
	Privacy is not simply an absence of information about us in the minds of others; rather it is the control we have over information about ourselves. \cite{CFried:Privacy}
\end{quote}
In order to realize this requirement in the modern IT world users are provided with user control functions, for example to decide about the access on their data for future use, and to view, modify, and delete unwanted data that had been collected. In its fundamental decision of May 13, 2014, the European Court has convicted Google to accept user demands to delete links to incorrect or irrelevant personal data in the Internet (C-131/12 – Google Spain SL). Such functions would constitute a so-called self-data-protection. It requires users to be highly aware of their rights and how to use them. Complementary, system-data-protection puts the load of data protection enforcement on the data collectors, ideally even without bothering the users. Typical system data protection functions are the deletion of personal data that are not bound to service purpose, the anonymization of personal data for research, and the abstinence from forwarding personal data to third parties. Together, system-data-protection and self-data-protection are supposed to provide a fair balance between the interests of service providers and their users.
In our research work we focus on self-data-protection according to Fried’s user control requirement \cite{CFried:Privacy}. A system data protection point of view would lead to other results, namely to a set of obligations for service providers. This is subject to further research.

@@

The first step is the \emph{world view} where all components are described in their current state. It consists of the following components: Assets, IT-Systems, Actors, Conflicts of Interests, Vulnerabilities, and Interactions.

The relevant \textbf{IT-Systems} were already described in Section \ref{subsubsection:it-systems}.
\textbf{Interactions} between assets, IT-Systems, humans, and vulnerabilites as partly described in Section \ref{subsubsection:it-systems} and will be further analyzed in Section \ref{subsec:pot-analysis}.

@@

In this scenario we focus our attention to only one asset: The privacy of the citizen. Our definition of privacy is described in detail in Section \ref{sec:privacydef}.

@@

\subsection{Step 2. Potential World Analysis}
\label{subsec:pot-analysis}

The \emph{Potential World Analysis} displays the intended and unintended interactions of the components in the world view.

The intended interactions support the underlying business objectives. Unintended interactions can lead to threats. Obviously this provides a conflict of interest with the victim's business model, to keep the asset away from unauthorized access. From the point of view of the victim, an attack would be an unintended interaction.

The potential world view consists of the following components: Business Objectives, Threats, hances/Risk, and Security Requirements. 
\textbf{Business Objectives} of the system owner were already described in Section \ref{subsubsection:it-systems} and \ref{subsec:world-analysis}. \textbf{Security Requirement}s are included to our analysis in Section \ref{subsec:priv-rec}.

\subsubsection{Threats}

A threat is a potential interaction of the components that targets an asset.
We restrict ourselves to the case of attacks, and the asset of privacy.
An attack is an interaction that is executed by an actor in response to a conflict of interest by exploiting a vulnerability of the system. We describe threats for the citizen privacy in the system.
This list can necessarily not be complete, but we make a best effort to cover the most relevant cases.

\textbf{T1. Insufficient Control Features.}
As soon as collected data of Citizens is stored on System Provider servers, all control over that data is lost. Although, control capabilities for Citizens add to the system complexity, which could motivate to omit those. Therefore interests of Citizens and System Providers are in conflict, namely it is the Citizen's \textit{Privacy vs. System Complexity} for System Providers.

\textbf{T2. Excessive Data Mining.}
The System Provider and/or the Local Authority secretly extract more private information from the collected data, than the Citizen agreed to. But results of the mining process create no disadvantages for Citizens, because there is no disclosure to third parties. This could be the case for a System Provider, who wants to test a new product and uses the pre-existing data collection.

Thus there are two possible conflicts: \textit{Privacy vs. Financial Profit of Serivce Provider} and  \textit{Privacy vs. Buisness Intelligence of Local Authority}
The threat can be provoked by either lax data handling policies of both System Providers and Local Authorities, or a weak law enforcement of existing supervision.
But the main issue, which can lead to such threats, is again a general \textit{Missing Privacy Awareness}.
